\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-12-10 paper.qmd
working on curiosities section: look for DHC on things that need fixing

## A curious difference between tSNE, UMAP and PaCMAP revealer

@@

In this section, the evaluation focuses on how effectively the tSNE, UMAP, and PaCMAP algorithms preserve the clustering structure. The data consists of five $4\text{-}D$ clusters, each with $1000$ points generated from a multivariate normal distribution with a small variance of $0.0025$. The first cluster is centered at $(0, 0, 0, 0)$, the second at $(1, 0, 0, 0)$, the third at $(0, 1, 0, 0)$, the fourth at $(0, 0, 1, 0)$, and the fifth at $(0, 0, 0, 1)$.

@@

In the tSNE layout (@fig-five-gau-projs (a1)), the distances between clusters appear smaller, leading to more tightly packed clusters, which is typical of tSNE's focus on preserving local structures. This often results in less emphasis on the global distances between clusters. On the other hand, UMAP (@fig-five-gau-projs (b1)) tends to maintain larger separations between clusters, better reflecting the global structure and preserving the relative distances between clusters more effectively. In contrast, PaCMAP (@fig-five-gau-projs (c1)) aim to balance these aspects by maintaining the distance between clusters and accurately positioning them, effectively capturing both global and local structures.

By visualizing the model of different layouts in $4\text{-}D$ reveal different characteristics of each NLDR method. When examining the model of a specific cluster, which appears more like a filled-out shape in tSNE (@fig-five-gau-projs (a2, a3)), and UMAP (@fig-five-gau-projs (b2, b3)), while the model has a pancake shape in PaCMAP (@fig-five-gau-projs (c2, c3)). The reason to have filled-out shaped clusters in tSNE and UMAP is that the ability of capture the local structure. On the other hand, the reason for the pancake shape in PaCMAP is the fail to capture the local structure.

@@

The differences in performance can be attributed to several factors. One key element is the choice of hyper-parameters, particularly the number of neighbors considered during the dimensionality reduction process. This parameter significantly impacts how well local structures are maintained. Additionally, the balance between global and local structures, and issues related to optimization and initialization all contribute to the observed outcomes.

It's important to note that these layouts are made with default hyper-parameter settings. Adjusting these parameters could lead to different results, potentially improving PaCMAP's ability to capture local structures or altering how tSNE and UMAP handle global and local structures. Thus, while PaCMAP generally balances global and local preservation, careful tuning of its parameters is important for achieving the best results in any given dataset.

@@

tsne_gau <- tsne_gau_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.3, color = "#000000") +
  geom_rect(aes(xmin = -0.05, xmax = 0.35, ymin = 0.28, ymax = 0.6),
               fill = "transparent", color = "red", linewidth = 0.8)

## Compute hexbin parameters
num_bins_x_gau1 <- 13
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

error_df_gau_tsne <- error_df_gau_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.5) + 
  xlim(c(0.3, 0.66)) + ylim(c(-0.01, 0.35)) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("a3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_segment(data = tr_from_to_df_gau1, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2,
    #size = 0.1,
    color = "#999999"
  ) + 
  geom_rect(
      aes(xmin = 0.3, 
          xmax = 0.65, 
          ymin = -0.05, 
          ymax = 0.35),
          fill = "transparent", 
      color = "red", 
      linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

@@

error_df_gau_umap <- error_df_gau_abs |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.5) + 
  xlim(c(0.47, 0.5475)) + ylim(c(0.0125, 0.09)) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("b3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_segment(data = tr_from_to_df_gau1, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.05,
    #size = 0.1,
    color = "#999999"
  ) + 
  geom_rect(
    aes(xmin = 0.45, 
        xmax = 0.57, 
        ymin = 0, 
        ymax = 0.1),
    fill = "transparent", 
    color = "red", 
    linewidth = 0.5)  +
  interior_annotation("b1", 
                      position = c(0.06, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

error_df_gau_pacmap <- error_df_gau_abs |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.6,
             size = 1.5) + 
  xlim(c(-0.02, 0.18)) + ylim(c(0.12, 0.32)) +
  scale_colour_continuous_sequential(palette = "YlOrRd", breaks = 20) +
  interior_annotation("c3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_segment(data = tr_from_to_df_gau1, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2
    ),
    alpha = 0.3,
    #size = 0.1,
    color = "#999999"
  ) +
  geom_rect(
    aes(xmin = -0.04, 
        xmax = 0.22, 
        ymin = 0.08, 
        ymax = 0.34),
    fill = "transparent", 
    color = "red", 
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

>>>>

## Curiosities

@@

With the drawing of the model in the data, several interesting differences between NLDR methods can be observed.

### Ordering of points

DHC: Will need to have more bins in each of the methods to illustrate this properly.

To illustrate a difference in how the methods organise points in the NLDR layout, simulated 4D data having five Gaussian clusters is used. @fig-five-gau-projs a1, b1, c1 show the 2D layouts for (a) t-SNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters are used. All three methods show the five clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the 4D space. In this projection the difference between methods can be seen. These clusters are fully 4D in nature, so we would expect the model to be a *crumpled 2D sheet* that stretches in all four dimensions. This is what is observed for t-SNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the 2D PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in 4D doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model fit (@fig-five-gau-projs a3, b3, c3). With t-SNE and UMAP the error is relatively uniformly distributed, but with PaCMAP there is more error in the center of the 2D layout, reflecting that there are more points in the middle that are further from the fitted model.

<!-- In this section, the evaluation focuses on how effectively the tSNE, UMAP, and PaCMAP algorithms preserve the clustering structure. The data consists of five $4\text{-}D$ clusters, each with $1000$ points generated from a multivariate normal distribution with a small variance of $0.0025$. The first cluster is centered at $(0, 0, 0, 0)$, the second at $(1, 0, 0, 0)$, the third at $(0, 1, 0, 0)$, the fourth at $(0, 0, 1, 0)$, and the fifth at $(0, 0, 0, 1)$.-->

@@

<!-- In the tSNE layout (@fig-five-gau-projs (a1)), the distances between clusters appear smaller, leading to more tightly packed clusters, which is typical of tSNE's focus on preserving local structures. This often results in less emphasis on the global distances between clusters. On the other hand, UMAP (@fig-five-gau-projs (b1)) tends to maintain larger separations between clusters, better reflecting the global structure and preserving the relative distances between clusters more effectively. In contrast, PaCMAP (@fig-five-gau-projs (c1)) aim to balance these aspects by maintaining the distance between clusters and accurately positioning them, effectively capturing both global and local structures.-->

<!-- By visualizing the model of different layouts in $4\text{-}D$ reveal different characteristics of each NLDR method. When examining the model of a specific cluster, which appears more like a filled-out shape in tSNE (@fig-five-gau-projs (a2, a3)), and UMAP (@fig-five-gau-projs (b2, b3)), while the model has a pancake shape in PaCMAP (@fig-five-gau-projs (c2, c3)). The reason to have filled-out shaped clusters in tSNE and UMAP is that the ability of capture the local structure. On the other hand, the reason for the pancake shape in PaCMAP is the fail to capture the local structure. -->

@@

<!-- The differences in performance can be attributed to several factors. One key element is the choice of hyper-parameters, particularly the number of neighbors considered during the dimensionality reduction process. This parameter significantly impacts how well local structures are maintained. Additionally, the balance between global and local structures, and issues related to optimization and initialization all contribute to the observed outcomes.

It's important to note that these layouts are made with default hyper-parameter settings. Adjusting these parameters could lead to different results, potentially improving PaCMAP's ability to capture local structures or altering how tSNE and UMAP handle global and local structures. Thus, while PaCMAP generally balances global and local preservation, careful tuning of its parameters is important for achieving the best results in any given dataset.-->

@@

tsne_gau <- tsne_gau_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.3) +
  geom_rect(aes(xmin = -0.05, xmax = 0.35, ymin = 0.28, ymax = 0.6),
               fill = "transparent", color = "grey50", linewidth = 0.8)

## Compute hexbin parameters
num_bins_x_gau1 <- 20 #13
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

error_df_gau_tsne <- error_df_gau_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) +  
  xlim(c(0.3, 0.66)) + ylim(c(-0.01, 0.35)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("a3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_tsne <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2) + 
  geom_rect(
      aes(xmin = 0.3, 
          xmax = 0.65, 
          ymin = -0.05, 
          ymax = 0.35),
          fill = "transparent", 
      color = "grey50", 
      linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

# library(langevitour)

@@

error_df_gau_umap <- error_df_gau_abs |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) +  
  xlim(c(0.47, 0.5475)) + ylim(c(0.0125, 0.09)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("b3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_umap <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.05) + 
  geom_rect(
    aes(xmin = 0.45, 
        xmax = 0.57, 
        ymin = 0, 
        ymax = 0.1),
    fill = "transparent", 
    color = "grey50", 
    linewidth = 0.5)  +
  interior_annotation("b1", 
                      position = c(0.06, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

error_df_gau_pacmap <- error_df_gau_abs |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) + 
  xlim(c(-0.02, 0.18)) + ylim(c(0.12, 0.32)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", breaks = 20) +
  interior_annotation("c3", 
                      position = c(0.08, 0.9),
                      cex = 2)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2
    ),
    alpha = 0.3) +
  geom_rect(
    aes(xmin = -0.04, 
        xmax = 0.22, 
        ymin = 0.08, 
        ymax = 0.34),
    fill = "transparent", 
    color = "grey50", 
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

### The effect of density

### 2024-12-10 paper.tex
working on curiosities section: look for DHC on things that need fixing

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-NLDR-variety-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-two-curvy-true-proj-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-NLDR-two-curvy-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/best-fit-umap1-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-bins-two-curvy-1.pdf}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-param-two-curvy-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-rm-lg-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.6\textheight]{paper_files/figure-pdf/fig-two_non_linear_diff_shaped_close_clusters-mse-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{paper_files/figure-pdf/best-fit-umap-tsne-pacamp-1.pdf}

@@

\section{A curious difference between tSNE, UMAP and PaCMAP
revealer}\label{a-curious-difference-between-tsne-umap-and-pacmap-revealer}

In this section, the evaluation focuses on how effectively the tSNE,
UMAP, and PaCMAP algorithms preserve the clustering structure. The data
consists of five \(4\text{-}D\) clusters, each with \(1000\) points
generated from a multivariate normal distribution with a small variance
of \(0.0025\). The first cluster is centered at \((0, 0, 0, 0)\), the
second at \((1, 0, 0, 0)\), the third at \((0, 1, 0, 0)\), the fourth at
\((0, 0, 1, 0)\), and the fifth at \((0, 0, 0, 1)\).

In the tSNE layout (Figure~\ref{fig-five-gau-projs} (a1)), the distances
between clusters appear smaller, leading to more tightly packed
clusters, which is typical of tSNE's focus on preserving local
structures. This often results in less emphasis on the global distances
between clusters. On the other hand, UMAP
(Figure~\ref{fig-five-gau-projs} (b1)) tends to maintain larger
separations between clusters, better reflecting the global structure and
preserving the relative distances between clusters more effectively. In
contrast, PaCMAP (Figure~\ref{fig-five-gau-projs} (c1)) aim to balance
these aspects by maintaining the distance between clusters and
accurately positioning them, effectively capturing both global and local
structures.

By visualizing the model of different layouts in \(4\text{-}D\) reveal
different characteristics of each NLDR method. When examining the model
of a specific cluster, which appears more like a filled-out shape in
tSNE (Figure~\ref{fig-five-gau-projs} (a2, a3)), and UMAP
(Figure~\ref{fig-five-gau-projs} (b2, b3)), while the model has a
pancake shape in PaCMAP (Figure~\ref{fig-five-gau-projs} (c2, c3)). The
reason to have filled-out shaped clusters in tSNE and UMAP is that the
ability of capture the local structure. On the other hand, the reason
for the pancake shape in PaCMAP is the fail to capture the local
structure.

The differences in performance can be attributed to several factors. One
key element is the choice of hyper-parameters, particularly the number
of neighbors considered during the dimensionality reduction process.
This parameter significantly impacts how well local structures are
maintained. Additionally, the balance between global and local
structures, and issues related to optimization and initialization all
contribute to the observed outcomes.

It's important to note that these layouts are made with default
hyper-parameter settings. Adjusting these parameters could lead to
different results, potentially improving PaCMAP's ability to capture
local structures or altering how tSNE and UMAP handle global and local
structures. Thus, while PaCMAP generally balances global and local
preservation, careful tuning of its parameters is important for
achieving the best results in any given dataset.

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-five-gau-projs-1.pdf}

@@

@@

\includegraphics[width=1\textwidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-1.pdf}

@@

\includegraphics[width=1\textwidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

@@

\includegraphics[width=1\textwidth,height=1\textheight]{paper_files/figure-pdf/fig-mnist-mse-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-mnist-tri-proj-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-error-mnist-1.pdf}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-NLDR-variety-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-two-curvy-true-proj-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-NLDR-two-curvy-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.3\textheight]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

{\centering \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/best-fit-umap1-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.3\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-bins-two-curvy-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-param-two-curvy-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-rm-lg-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.6\textheight]{paper_files/figure-pdf/fig-two_non_linear_diff_shaped_close_clusters-mse-1.pdf}

@@

{\centering \includegraphics[width=1\linewidth,height=0.9\textheight]{paper_files/figure-pdf/best-fit-umap-tsne-pacamp-1.pdf}

@@

\section{Curiosities}\label{curiosities}

With the drawing of the model in the data, several interesting
differences between NLDR methods can be observed.

\subsection{Ordering of points}\label{ordering-of-points}

DHC: Will need to have more bins in each of the methods to illustrate
this properly.

To illustrate a difference in how the methods organise points in the
NLDR layout, simulated 4D data having five Gaussian clusters is used.
Figure~\ref{fig-five-gau-projs} a1, b1, c1 show the 2D layouts for (a)
t-SNE, (b) UMAP, and (c) PaCMAP, respectively. The default
hyper-parameters are used. All three methods show the five clusters,
with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster
to illustrate the curious detail. Figure~\ref{fig-five-gau-projs} a2,
b2, c2 show the fitted models in a projection of the 4D space. In this
projection the difference between methods can be seen. These clusters
are fully 4D in nature, so we would expect the model to be a
\emph{crumpled 2D sheet} that stretches in all four dimensions. This is
what is observed for t-SNE and UMAP. The curious detail is that the
model for PaCMAP is closer to a \emph{pancake} in shape! What this means
is that there has to be some ordering of points in the 2D PaCMAP layout
that induces the flat model, likely that the points are organised by the
global principal components. So the layout of the model in 4D doesn't
reflect the dimension of each cluster. DHC: is there something in the
documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model
fit (Figure~\ref{fig-five-gau-projs} a3, b3, c3). With t-SNE and UMAP
the error is relatively uniformly distributed, but with PaCMAP there is
more error in the center of the 2D layout, reflecting that there are
more points in the middle that are further from the fitted model.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-five-gau-projs-1.pdf}

@@

\subsection{The effect of density}\label{the-effect-of-density}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-1.pdf}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-mnist-mse-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-mnist-tri-proj-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-error-mnist-1.pdf}
